---
title: "Time series analysis"
output: html_notebook
---

This notebook follows this  [tutorial](https://nwfsc-timeseries.github.io/atsa-labs/chap-tslab.html).

```{r}
install.packages('stats')
install.packages('MARSS')
install.packages('forecast')
install.packages('datasets')
install.packages("tidyverse")
```

```{r}
install.packages('devtools')

```
```{r}
install.packages("rmarkdown")
library(rmarkdown)
```
# Linear regression in matrix form

```{r}
data("stackloss", package = "datasets")
dat= stackloss[1:4,]
dat

```

## simple regression:
```{r}
#regress stack loss against air flow
fit = lm(stack.loss ~ Air.Flow, data =dat)
Z = model.matrix(fit)
Z[1:4,]
```

```{r}
#using the matrix COVARIANCE formula
y= matrix(dat$stack.loss, ncol=1)
Z = cbind(1, dat$Air.Flow)
solve(t(Z) %*% Z) %*% t(Z) %*% y
```


```{r}
coef(fit)
```
```{r}
#now with multiple explanatory variables:
fit_m = lm(stack.loss ~ Air.Flow + Water.Temp + Acid.Conc., data=dat)
```


```{r}
Z = model.matrix(fit_m)
Z
```

```{r}
#now using matrices:
y = matrix(dat$stack.loss, ncol=1)
Z = cbind(1, dat$Air.Flow, dat$Water.Temp, dat$Acid.Conc.)
solve(t(Z) %*% Z) %*% t(Z) %*% y
```


```{r}
coef(fit_m)
```


```{r}
#now use the transform of form 1
y= matrix(dat$stack.loss, nrow=1)
d=rbind(1, dat$Air.Flow, dat$Water.Temp, dat$Acid.Conc.)
y %*% t(d) %*% solve(d %*% t(d))
```
## Adding a seasonal effect to the model

```{r}
#adding in a seasonal effect

dat = cbind(dat, qtr = paste(rep("qtr", 4), 1:4, sep = ""))
dat
```

```{r}
#fit the model with lm

fit_s = lm(stack.loss ~ -1+qtr, data=dat)#with the -1 we get rid of the intercept
coef(fit_s)
```
## Adding a seasonal effect + other explanatory variables

```{r}
data(stackloss, package = "datasets")
fulldat = stackloss
n = nrow(fulldat)
fulldat = cbind(fulldat, owner = rep(c("sue", "aneesh", "joe"), 
    n)[1:n], qtr = paste("qtr", rep(1:4, n)[1:n], sep = ""), 
    reg = rep(c("n", "s"), n)[1:n])
```

```{r}
#lets fit the model:
fit_s1 = lm(stack.loss ~ -1 + qtr + Air.Flow:qtr:owner, data = fulldat)
Z = model.matrix(fit_s1)
```

```{r}
#confounding parameters:
f = lm(stack.loss ~ -1 + Air.Flow + reg + qtr, data = fulldat)
coef(f)
```
```{r}
model.matrix(f)
```

# examples of time series

```{r}
data("WWWusage")
par(mai=c(0.9, 0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(WWWusage, ylab = "", las =1, col = 'blue', lwd=2)
```

```{r}
lx <- log(AirPassengers)
par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 0, 0, 0))
plot.ts(lx, las = 1, ylab = "")
```

```{r}
library(devtools)
devtools::install_github("r-lib/svglite")
install.packages("kableExtra")
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS=TRUE)
devtools::install_github("nwfsc-timeseries/atsar")
devtools::install_github("nwfsc-timeseries/atsalibrary")
```



```{r}
data(NHTemp, package = "atsalibrary")
Temp <- NHTemp
data(MLCO2, package = "atsalibrary")
CO2 <- MLCO2
data(hourlyphyto, package = "atsalibrary")
phyto_dat <- hourlyphyto
```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

